A Way Forward in Parallelising Dynamic Languages

### 2014-05-05 talk/icooolps2014/position-paper.tex
at least give HTM a 0 too
emphasize atomic block parallelisation

These requirements are not easy to meet. We argue that STM is the
overall winner. While it has a big performance problem currently, it
gets more points in all the other categories. We think that it is the
only solution that also provides a better synchronisation mechanism to
the application in the form of atomic blocks.

@@

\begin{table*}[ht]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    & \textbf{GIL} & \textbf{Fine-grained locking}
    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
    \hline
    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
    \hline
    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
    \hline
    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
    \hline
    Better synchronisation        & o    & o  & o    & -  & ++   \\
    \hline
    Implementation                & ++   & -  & ++   & ++ & ++   \\
    \hline
    External libraries            & ++   & ++ & ++   & ++ & ++   \\
    \hline
  \end{tabular}
  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
  \label{tab:comparison}
\end{table*}

@@

Since the GIL is mostly an implementation detail of the interpreter,
it is not exposed to the application running on top of it. To
synchronise memory accesses in applications using threads, the
state-of-the-art still means explicit locking everywhere. It is well
known that using locks for synchronisation is not
easy~\cite{christopher10,victor11,shan08}. They are non-composable,
have overhead, may deadlock, limit scalability, and add to the overall
complexity of the program logic. For a better parallel programming
model for dynamic languages, we propose another, well-known
synchronisation mechanism called \emph{atomic
  blocks}~\cite{tim03,tim05}.

Atomic blocks are composable, deadlock-free, higher-level and expose
useful atomicity and isolation guarantees to the application for a
series of instructions.  Interpreters using a GIL can simply guarantee
that the GIL is not released during the execution of the atomic
block. Of course, this still means that no two atomic blocks can
execute in parallel or even concurrently. Potential solutions that
provide a good way to implement atomic blocks are therefore
preferable.

@@

\begin{description}
\item[Performance:] How well does the approach perform compared to the
  GIL on a single and on multiple threads?
\item[Existing applications:] How big are the changes required to
  integrate with and parallelise existing applications?
\item[Better synchronisation:] Does the approach enable better
  synchronisation mechanisms for applications (e.g. atomic blocks)?
\item[Implementation:] How difficult is it to implement the approach
  in the interpreter?
\item[External libraries:] Does the approach allow for easy
  integration of external libraries?
\end{description}

@@

The performance of HTM is pretty good as it does not introduce much
overhead ($<40\%$ overhead~\cite{odaira14}). And it can transparently
parallelise existing applications to some degree. The implementation
is very straight-forward because it directly replaces the GIL in a
central place. HTM is also directly compatible with any external
library that needs to be integrated and synchronised for use in
multiple threads. The one thing that is missing is support for a
better synchronisation mechanism for the application. It is not
possible in general to expose the hardware-transactions to the
application in the form of atomic blocks because that would require
much longer transactions.

>>>>

These requirements are not easy to meet. We argue that STM is the
overall winner. While it has a big performance problem currently, it
gets more points in all the other categories. We think that it is the
only solution that also provides a better synchronisation mechanism to
the application in the form of paralleliseable atomic blocks.

@@

\begin{table*}[ht]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    & \textbf{GIL} & \textbf{Fine-grained locking}
    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
    \hline
    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
    \hline
    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
    \hline
    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
    \hline
    Better synchronisation        & o    & o  & o    & o  & ++   \\
    \hline
    Implementation                & ++   & -  & ++   & ++ & ++   \\
    \hline
    External libraries            & ++   & ++ & ++   & ++ & ++   \\
    \hline
  \end{tabular}
  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
  \label{tab:comparison}
\end{table*}

@@

Since the GIL is mostly an implementation detail of the interpreter,
it is not exposed to the application running on top of it. To
synchronise memory accesses in applications using threads, the
state-of-the-art still means explicit locking everywhere. It is
known that using locks for synchronisation can be hard at
times~\cite{christopher10,victor11,shan08}. They are non-composable,
have overhead, may deadlock, limit scalability, and add to the overall
complexity of the program logic. For a better parallel programming
model for dynamic languages, we propose another, well-known
synchronisation mechanism called \emph{atomic
  blocks}~\cite{tim03,tim05}.

Atomic blocks are composable, deadlock-free, higher-level and expose
useful atomicity and isolation guarantees to the application for a
series of instructions. Interpreters using a GIL can simply guarantee
that the GIL is not released during the execution of the atomic
block. Of course, this still means that no two atomic blocks can
execute in parallel or even concurrently.  Potential solutions are
preferable if they provide a good way to implement atomic blocks that
are also able to be executed in parallel.

@@

\begin{description}
\item[Performance:] How well does the approach perform compared to the
  GIL on a single and on multiple threads?
\item[Existing applications:] How big are the changes required to
  integrate with and parallelise existing applications?
\item[Better synchronisation:] Does the approach enable better, paralleliseable
  synchronisation mechanisms for applications (e.g. atomic blocks)?
\item[Implementation:] How difficult is it to implement the approach
  in the interpreter?
\item[External libraries:] Does the approach allow for easy
  integration of external libraries?
\end{description}

@@

The performance of HTM is pretty good as it does not introduce much
overhead ($<40\%$ overhead~\cite{odaira14}). And it can transparently
parallelise existing applications to some degree. The implementation
is very straight-forward because it directly replaces the GIL in a
central place. HTM is also directly compatible with any external
library that needs to be integrated and synchronised for use in
multiple threads. The one thing that is missing is support for a
better synchronisation mechanism for the application. It is not
possible in general to expose the hardware-transactions to the
application in the form of atomic blocks, because that would require
much longer transactions.
